Survival in chessland

### 2018-12-16 trunk/chess/papers/survival.tex
Underline safest squares, paper text

@@

\title{Survival in chessland}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2018 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2018 with the threefold
repetition of the Association for Computational Heresy; 
{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
53 Centipawns} }

@@

  Of course, this all relies on the assumption that if
  chess-to-the-death ends in a draw, the players are spared or allowed
  to repeat indefinitely. If both players are actually executed, then
  this line is truly a Fool's Draw!
}
%
\footnotetext{
  But is it? The lichess implementation (although known
  to be buggy\cite{kingme}) does not permit a threefold repetition
  claim in this situation, which got me thinking that maybe there is
  some subtlety here. Is the starting position special somehow, not
  counting as having occurred? The relevant statute, from the FIDE
  Laws of Chess\cite{fiderules}:
  \begin{quote}
    {\bf 9.2}.\quad The game is drawn, upon a correct claim by a
    player having the move, when the same position for at least the
    third time (not necessarily by a repetition of moves):
    \begin{enumerate}[label=\alph*.]
    \item is about to appear, if he first writes his move, which
      cannot be changed, on his scoresheet and declares to the arbiter
      his intention to make this move, or
    \item has just appeared, and the player claiming the draw has the
      move.
  \end{enumerate}

  So the question is, has the starting position ``appeared'' before
  white's first move? The rules are not totally clear on this point.
  Note that ``positions are considered the same'' only when the
  same player ``has the move.'' FIDE defines ``have the move'' as
  \begin{quote}
    1.3.\quad A player is said to 'have the move' when his opponent's
    move has been 'made'.
  \end{quote}
  A strong case can therefore be made that white does not 'have the
  move' in the formal sense at the beginning of the game, since black
  has not made a move!

  Nonetheless, it does seem clear that white can claim a draw by
  9.2.a, by committing the move 5. Nf3 and declaring to the arbiter
  that the position is now {\it about to appear} for the third time.
  This seems unambiguously legal. }
%
But what if you are forced to {\it be one of the chess pieces} to the
death? That is, your little soul inhabits one of the 32 pieces or
pawns and your soul is vanquished if that piece is eliminated. Now it
doesn't matter whether you're good or bad at chess, because you don't
get to pick what happens in the game. What matters is that your piece
lives to the end of the game, when all surviving pieces are set free.
Which piece should you want to be?

@@

@@

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their black
counterparts. The \pawn e is vastly more dead than \Pawn e. Note
somewhat satisfyingly that the \pawn c has the highest variance; this
was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
  in the popular Queen's Gambit (1. d4 d5 2. c4); variance may come
  from individual players having a strong preference to accept or
  decline.\footnote{XXX actually this can't be true because we group
    by the white player, but the black player chooses what to do.}

% XXX

@@

>>>>

\interfootnotelinepenalty=0

@@

\title{Survival in chessland}
\author{Dr.~Tom~Murphy~VII~Ph.D.\footnote{
Copyright \copyright\ 2018 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2018 with the threefold
repetition of the Association for Computational Heresy; 
{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
53 Centipawns} }

@@

  Of course, this all relies on the assumption that if
  chess-to-the-death ends in a draw, the players are spared or allowed
  to repeat indefinitely. If both players are actually executed, then
  this line is truly a Fool's Draw!
}

But what if you are forced to {\it be one of the chess pieces} to the
death? That is, your little soul inhabits one of the 32 pieces or
pawns and your soul is vanquished if that piece is eliminated. Now it
doesn't matter whether you're good or bad at chess, because you don't
get to pick what happens in the game. What matters is that your piece
lives to the end of the game, when all surviving pieces are set free.
Which piece should you want to be?

@@

\footnotetext{
  But is it? The lichess implementation (although known
  to be buggy\cite{kingme}) does not permit a threefold repetition
  claim in this situation, which got me thinking that maybe there is
  some subtlety here. Is the starting position special somehow, not
  counting as having occurred? The relevant statute, from the FIDE
  Laws of Chess\cite{fiderules}:
  \begin{quote}
    {\bf 9.2}.\quad The game is drawn, upon a correct claim by a
    player having the move, when the same position for at least the
    third time (not necessarily by a repetition of moves):
    \begin{enumerate}[label=\alph*.]
    \item is about to appear, if he first writes his move, which
      cannot be changed, on his scoresheet and declares to the arbiter
      his intention to make this move, or
    \item has just appeared, and the player claiming the draw has the
      move.
  \end{enumerate}

  So the question is, has the starting position ``appeared'' before
  white's first move? The rules are not totally clear on this point.
  Note that ``positions are considered the same'' only when the
  same player ``has the move.'' FIDE defines ``have the move'' as
  \begin{enumerate}[label=\arabic*.3.] % hack to get 1.3.
    \item A player is said to 'have the move' when his opponent's
    move has been 'made'.
  \end{enumerate}
  A strong case can therefore be made that white does not 'have the
  move' in the formal sense at the beginning of the game, since black
  has not made a move!

  Nonetheless, it does seem clear that white can claim a draw by
  9.2.a, by committing the move 5. Nf3 and declaring to the arbiter
  that the position is now {\it about to appear} for the third time.
  This seems unambiguously legal. }

@@

% XXX: Move this stuff to the appendix.

@@

@@

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4); this may be a
good example of a piece that has very different survival rates in
different opening preferences. Since the variance is otherwise very
low, I did not attempt to

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%.

\section{Safest spaces}

The fate of each piece is to either survive or die, and it does so on
one of the 64 squares. With the same replay of the 500 million games
I also kept statistics on the fates of each piece. If being a chess
piece to the death, and possessing some influence over where your piece
moves, it may be helpful to know where to go. Even without influence,
such knowledge could help calibrate your anxiety.

Other than the bishops---which have no legal way to reach half of the
squares---every piece reaches every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

% There may even be hidden achievements for reaching these squares!

That said, there are characteristic patterns for each of the pieces,
which make sense given their starting positions and patterns of
movement. You could probably guess the piece just by looking at one of
the heat maps below, although---spoiler alert---the piece is just
listed right there and they are in order. Two independent things are
communicated in these graphics: The chance that the piece ends the
game on some square (alive or dead), and its survival chances there.
In each map, a darker background color indicates that the piece ends
on that square more often. The shade is based on the rank (64/64 black
is most common, 63/64 black is next most common, etc.) rather than the
absolute probability, since otherwise the graphic looks boring. The
number on the square is the percentage survival of the piece when it
is last seen (either being captured or surviving to the end of the
game) on that square. The four squares with the highest survival rates
are underlined for your convenience.

\newcommand{\chessmap}[2]{\begin{center} \includegraphics[width=0.65 \linewidth]{#1} {#2} \end{center}\vspace{-0.1in}}

\chessmap{piece0}{\Rook a8}
\chessmap{piece1}{\Knight b8}
\chessmap{piece2}{\Bishop c8}
\chessmap{piece3}{\Queen d8}
\chessmap{piece4}{\King e8}
\chessmap{piece5}{\Bishop f8}
\chessmap{piece6}{\Knight g8}
\chessmap{piece7}{\Rook h8}

\chessmap{piece8}{\Pawn a7}
\chessmap{piece9}{\Pawn b7}
\chessmap{piece10}{\Pawn c7}
\chessmap{piece11}{\Pawn d7}
\chessmap{piece12}{\Pawn e7}
\chessmap{piece13}{\Pawn f7}
\chessmap{piece14}{\Pawn g7}
\chessmap{piece15}{\Pawn h7}

\chessmap{piece16}{\pawn a2}
\chessmap{piece17}{\pawn b2}
\chessmap{piece18}{\pawn c2}
\chessmap{piece19}{\pawn d2}
\chessmap{piece20}{\pawn e2}
\chessmap{piece21}{\pawn f2}
\chessmap{piece22}{\pawn g2}
\chessmap{piece23}{\pawn h2}

\chessmap{piece24}{\rook a1}
\chessmap{piece25}{\knight b1}
\chessmap{piece26}{\bishop c1}
\chessmap{piece27}{\queen d1}
\chessmap{piece28}{\king e1}
\chessmap{piece29}{\bishop f1}
\chessmap{piece30}{\knight g1}
\chessmap{piece31}{\rook h1}
